Data Specialization for a Staged Lambda Calculus

### 2015-01-27 paper/splitting.tex
edits all over (syntax) and masking section

Given a term $e$ in the source language \lang, where terms of
stage~\bbone\ and stage~\bbtwo\ may be interleaved, \emph{(stage)
  splitting} separates the stage~\bbone\ and stage~\bbtwo\
subcomputations into two terms,~$e_1$ and~$e_2$, in an unstaged target language \langmono\ 
%each performing the
%subcomputations corresponding to each stage.  More precisely, given a
%\lang\ term~$e$, splitting yields two terms $e_1$ and $e_2$ in a
%unstaged target language \langmono\
such that~$e_1$ performs all the
stage~\bbone\ subcomputations in~$e$ and~$e_2$~performs all of the
stage~\bbtwo\ subcomputations. Splitting preserves correctness in the sense that evaluating~$e$ in
the staged language produces the same result as evaluating~$e_1$ in the target language,
then further evaluating~$e_2$ in the target language with the result from~$e_1$.

In this section, we present a
splitting algorithm for the source language \lang, specifying
precisely the algorithm and its correctness conditions.  
We begin by defining a masking relation that serves to extract stage~\bbone\ and
stage~\bbtwo\ components of \lang\ partial values into \langmono\ terms that the outputs of splitting are designed to match. We then 
move on to the specification of the splitting algorithm itself.

@@

Recall that partial values and residual tables produced by \bbone-evaluation may
contain content at \emph{both} stages. For example the term:
\begin{lstlisting}
1`(1+2, next{`2`7 + 5`1`})`
\end{lstlisting}
evaluates via $\redonesym$ to:
\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}
To match the results of \bbone-evaluation,
splitting the term should generate an~$e_1$ that reduces to the stage~\bbone\ content of the partial value, \texttt{3},
and an~$e_2$ corresponding to the stage~\bbtwo\ computation: \\
$\letin \yhat {\mathtt{7+5}} \yhat$.

to precisely specific the stage~\bbone\ and stage~\bbtwo\ components of \bbone-evaluation results, we define 
\emph{masking} functions, written $\masko{{-};{-}}$ and $\maskt{{-};{-}}$, 
which take a residual table and partial value as input and emit a term in \langmono\
that corresponds to only the content belonging to a single stage.
\ref{fig:valMask} defines these masking functions.

\emph{Stage~\bbone\ masking} of the partial value~$v$ (written $\masko{v}$) produces a term in \langmono\ containing exactly
the stage~\bbone\ content of~$v$. Since \next\ terms only contain stage~\bbtwo\
content, $\masko{v}$ replaces all $\next$ subterms of~$v$ with \texttt{()}.  Since lambdas may represent multi-stage computations, stage~\bbone\ masking splits the body of lambdas as a general stage~\bbone\ term (as described in \ref{sec:split-one}), and returns
the stage~\bbone\ component. $\masko{v}$ always produces a value in \langmono\ that has
the same structure as~$v$.  For example, the stage~\bbone\ mask of the partial value given above is \texttt{(3,())}.

\emph{Stage~\bbtwo\ masking} of the partial value $v$ and residual context~$\xi$ (written $\maskt{\xi;v}$) produces a term in \langmono\ containing
exactly the stage~\bbtwo\ content of $[\xi;v]$. Unlike stage~\bbone\ masking, stage~\bbtwo\ masking 
requires access to the residual table $\xi$ because after \bbone-evaluation, all stage~\bbtwo\
has been lifted out of~$v$. Stage~\bbtwo\ masking discards
stage~\bbone\ content in~$v$ by replacing base constants with~\texttt{()},
replacing $\next~\hat{y}$ with~$\hat{y}$, and replacing all $n$-ary term
constructors with an $n$-tuple of masked subterms. Like stage~\bbone\ masking, stage~\bbtwo\
masking splits the bodies of lambdas as general stage~\bbone\ terms, and
returns the stage~\bbtwo\ component. After discarding all stage~\bbone\ content from~$v$, stage~\bbtwo\ masking reifies $\xi$ around the result,
producing a term in \langmono. This term reduces to a value that with the same structure as~$v$.
For example, the stage~\bbtwo\ mask of the partial value and residual table given above is:
\begin{lstlisting}
let yhat=7+5 in ((),yhat)
\end{lstlisting}

@@

>>>>

\begin{concretesyntax}

The primary strength of \lang\ is the expressive power afforded by
its ability to syntactically interleave work intended for stage \bbone\
and work intended for stage \bbtwo, as well as it ability to form
mixed-stage abstractions.
[Sentence about how we actual want this in separated monostage form.]
In this section, we introduce a {\em splitting transformation} whose job 
is to translate \lang\ terms into a form where the stages are 
seperated {\em at the top level}.

Recall that $\colmix {e_1}{A_1}$ represents multistage computations with multistage results,
whereas $\coltwo {e_2}{A_2}$ represents multistage computions with stage \bbtwo\ results.
Similarly, our splitting algorithm has two forms, 
given by the judgements $\splitonesym$ and $\splittwosym$, 
which respectively take these types of \lang\ terms as input.
For the outputs, we introduce new constructs 
to encode multistage values and computations that are syntactically separated at the top level.

@@

Recall that partial values (including residual tables) produced by \bbone-evaluation may
contain content at \emph{both} stages. 
In particular, partial values have the character of a value at stage~\bbone\ and of a term at stage~\bbtwo.
For example, the term
\begin{lstlisting}
1`(1+2, next{`2`7 + 5`1`})`
\end{lstlisting}
evaluates via $\redonesym$ to:
\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}

What we desire is a simple way to translate this partial value into a form that separates the stage~\bbone\ content (the value \texttt{3}) 
from the stage~\bbtwo\ content (the term $\letin \yhat {\mathtt{7+5}} \yhat$),
allowing us to later define a term translation that respects this value translation.

For this purpose, we define a relation called {\em masking}, written $\vsplito$, 
which sends partial values to {\em masked values}.
Masked values have only one form, written $\mval {i} {r}$, 
where $i$ is a monostage value (the {\em immediate value}) and $r$ is a monostage term (the {\em resumer}).
The masking relation is defined in \ref{fig:valMask}.

Masking operates by first inducting on the entries of the residual table.  
Being purely stage~\bbtwo\ content, these are reified into let statements at the top of the resumer.
Once the table is empty, masking inducts on value itself.

Masking assigns purely stage~\bbone\ values to the immediate value
and likewise assign references into the residual table to the resumer.
In both cases, the offside component is assigned to \texttt{()}, to represent trivial information.

Masking distributes into tuples, injections, and rolls, since their subvalues may have content at both stages.
However, the tags of injections and rolls are replicated only in the immediate value, 
since they represent stage~\bbone\ information.

Since lambdas may represent multi-stage computations, 
masking splits the body of lambdas as a general stage~\bbone\ terms (as described in \ref{sec:split-one}), 
and packages the resulting terms as functions.

The exmaple above is masked into,
\begin{lstlisting}
mv((3,()), let yhat=7+5 in ((),yhat))
\end{lstlisting}

@@

\end{concretesyntax}
